---
title: "BW_clean"
output: html_document
---


---
title: "BW_random"
output: html_document
---

# load packages and data
```{r}
pacman::p_load(
  tidyverse,
  readxl,
  terra,
  stats,
  dismo,
  usdm,
  corrplot,
  PerformanceAnalytics,
  oddsratio,
  DHARMa,
  MASS,
  nlme,
  mgcv,
  mgcViz,
  bibtex)

write.bib(c("tidyverse",
  "readxl",
  "terra",
  "stats",
  "dismo",
  "usdm",
  "corrplot",
  "PerformanceAnalytics",
  "oddsratio",
  "DHARMa",
  "MASS",
  "nlme",
  "mgcv",
  "mgcViz",
  "bibtex"), file = "Rpackages.bib")


setwd("~/Documents/Master's/Quantitative Biology/BW_proj")
getwd()
env_r_all_na <- read_excel(path = "./Zhang2022RasterDataLayers/random_pt_xy.xlsx")
env_d_all_na <- read_excel(path = "./Zhang2022RasterDataLayers/defol_pt_xy.xlsx")
```

# prep data
### remove NAs
```{r echo=TRUE}
env_r_all <- env_r_all_na[complete.cases(env_r_all_na), ]
env_d_all <- env_d_all_na[complete.cases(env_d_all_na), ]
```

### create same number of samples for each data set
chose 600 bc largest common centennial 
```{r}
env_r <- env_r_all[sample(nrow(env_r_all), 2400), ]
env_d <- env_d_all[sample(nrow(env_d_all), 2400), ]
```

### combine onto one dataframe
```{r}
# make column specifying defoliation points
env_r$r_d <- c(rep(1, length(env_r$pcp0708)))

# make column specifying random points
env_d$r_d <- c(rep(0, length(env_d$pcp0708)))


# combine and drop columns not common to both and environmental variables not using
bw_env <- rbind(env_d[intersect(names(env_d),names(env_r))],env_r[intersect(names(env_d),names(env_r))])
bw_env <- bw_env[, -c(3, 5, 18, 21)]
# make tree type (SpRasterNA) categorical
bw_env<- filter(bw_env, SpRasterNA != 4)
bw_env$SpRasterNA <- as.factor(bw_env$SpRasterNA)

```

# correlation analysis
```{r}
##check Multicollinearity - between 14 predictors
predictors_num <- bw_env[, -c(3,18)]
# correlation structure
varCor <- cor(predictors_num)
corrplot(varCor,  tl.col='black')
chart.Correlation(varCor, histogram=F, pch=19)


# clustering with dendrogram
allDistNew <- abs(as.dist(cor(predictors_num)))
allClusNew <- hclust(1 - allDistNew)
plot(allClusNew, hang=-1)

# Variance Inflation Factor
usdm::vif(as.data.frame(predictors_num))
```

### remove variables
```{r}
# keep MaxT0506 and pcp0607 bc biologically relevant
## remove all but one MaxT (0506), remove sg_date and sg__gdd_p2, minT0607 and minT0708
predictors_num_sel <- predictors_num[,-c(3,4,9,10,12,13)]

# retry
## correlation structure
varCor_sel <- cor(predictors_num_sel)
corrplot(varCor_sel,  tl.col='black')
chart.Correlation(varCor_sel, histogram=F, pch=19)

## clustering with dendrogram
allDistNew_sel <- abs(as.dist(cor(predictors_num_sel)))
allClusNew_sel <- hclust(1 - allDistNew_sel)
plot(allClusNew_sel, hang=-1)

## VIF
usdm::vif(as.data.frame(predictors_num_sel))

## pcps still very correlated so maybe remove next highest which is pcp0708 and is correlated with X, also remove pcp0506 bc correlated with Y, keeping pcp0607 makes sense beyond biological pretense because the least with lat/long

# retry #2
predictors_num_sel2 <- predictors_num_sel[,-c(4,6)]

## correlation structure
varCor_sel2 <- cor(predictors_num_sel2)
corrplot(varCor_sel2, tl.col = "black")
chart.Correlation(varCor_sel2, histogram=F, pch=19)

## clustering with dendrogram
allDistNew_sel2 <- abs(as.dist(cor(predictors_num_sel2)))
allClusNew_sel2 <- hclust(1 - allDistNew_sel2)
plot(allClusNew_sel2, hang=-1)

## VIF
usdm::vif(as.data.frame(predictors_num_sel2))

pred_final <- predictors_num_sel2
```

# create data frame with final variables, tree species, and x/y coordinateas
```{r}
final_names <- c(names(pred_final), "r_d", "SpRasterNA")
str(final_names)
bw_env_f <- bw_env[ ,names(bw_env) %in% final_names]
```

```{r}
final_names_xy <- c(names(pred_final), "r_d", "SpRasterNA", "X", "Y")
str(final_names_xy)
bw_env_xy <- bw_env[ ,names(bw_env) %in% final_names_xy]
```

# GLM
```{r echo=TRUE}
m_1 <- glm(r_d ~  MaxT0506 + MinT0506 + pcp0607 + sg__gdd_p1 + DistanceRo + DistanceRi + SpRasterNA - 1, data = bw_env_xy, family = binomial(link = "logit"))
summary(m_1)

## check if type 3 - ctrl for each level of variable not just variable above by including variables in different order
 m_test1 <- glm(r_d ~  MinT0506 + pcp0607 +MaxT0506 - 1, data = bw_env_xy, family = binomial(link = "logit"))
summary(m_test1)
 m_test2 <- glm(r_d ~ MaxT0506+ pcp0607 +MinT0506 - 1, data = bw_env_xy, family = binomial(link = "logit"))
summary(m_test2)
```

## plot residuals against fitted
about DHARMa:
"The ‘DHARMa’ package uses a simulation-based approach to create readily interpretable scaled (quantile) residuals for fitted (generalized) linear mixed models.
The resulting residuals are standardized to values between 0 and 1 and can be interpreted as intuitively as residuals from a linear regression. The package also provides a number of plot and test functions for typical model misspecification problems, such as over/underdispersion, zero-inflation, and residual spatial and temporal autocorrelation.

To interpret the residuals, remember that a scaled residual value of 0.5 means that half of the simulated data are higher than the observed value, and half of them lower. A value of 0.99 would mean that nearly all simulated data are lower than the observed value. The minimum/maximum values for the residuals are 0 and 1. For a correctly specified model we would expect asymptotically a uniform (flat) distribution of the scaled residuals & uniformity in y direction if we plot against any predictor."
(Hartig 2022)
```{r echo=TRUE}
sim_res <- simulateResiduals(fittedModel = m_1, plot = F)
plot(sim_res, quantreg=T)   # normal along Q-Q plot
#red quantile lines mean significant deviation from straight line
jpeg("QQ-plot.jpg", width = 11, height =11, units = "cm", res = 1200, pointsize = 12)
plotQQunif(simulationOutput = sim_res, 
           testDispersion = FALSE,
           testUniformity = FALSE,
           testOutliers = FALSE)
dev.off()

jpeg("res-fitted plot.jpg", width = 11, height =11, units = "cm", res = 1200, pointsize = 12)
plotResiduals(sim_res, quantreg = T, colour = "black")
dev.off()


# check for outliers
outliers(sim_res)

# test to see where pattern is coming from
plotResiduals(sim_res, bw_env_f$sg__gdd_p1)
# sg_gdd_p1 good
plotResiduals(sim_res, bw_env_f$pcp0607)
# pcp0607 good
plotResiduals(sim_res, bw_env_f$MinT0506)
# MinT0708 quite bad
plotResiduals(sim_res, bw_env_f$MaxT0506)
# MaxT0506 good
plotResiduals(sim_res, bw_env_f$DistanceRo)
# DistanceRo not great fror 0.5 and 0.75
plotResiduals(sim_res, bw_env_f$DistanceRi)
# DistanceRi good
plotResiduals(sim_res, bw_env_f$SpRasterNA, asFactor = T)
# SpRasterNA good

# test autocorrelation
testSpatialAutocorrelation(simulationOutput = sim_res, x = bw_env_xy$X, y= bw_env_xy$Y)
# very significant
# plot against locations
plotResiduals(sim_res, bw_env_xy$X)
plotResiduals(sim_res, bw_env_xy$Y)

# residuals aren't homogenous

# plot against lag
acf(m_1$residuals, lag.max = 5000)
## pattern in the residuals  (likely spatial autocorrelation)
jpeg("lagplot glm.jpg", width = 11, height =10, units = "cm", res = 1200, pointsize = 12)
acf(m_1$residuals, lag.max = 5000, main = " ")
dev.off()


# these variables have pattern in similarity
acf(bw_env_xy$pcp0607, lag.max = 5000)
acf(bw_env_f$MaxT0506, lag.max = 5000)
acf(bw_env_xy$DistanceRo, lag.max = 5000)

# these variables are mediocre
acf(bw_env_xy$DistanceRi, lag.max = 5000)
acf(bw_env_xy$MinT0506, lag.max = 5000)

# these are fine
acf(bw_env_xy$sg__gdd_p1, lag.max = 5000)
acf(bw_env_xy$X, lag.max = 5000)
acf(bw_env_xy$Y, lag.max = 5000)
```

# comapre GLMs 
models to test
full model:
r_d ~ MaxT0506 + MinT0506 + pcp0607 + Sg__gdd_p1 + DistanceRo + DistanceRi + SpRasterNA
reduced model:
r_d ~ prp0607 + MaxT0506 + MinT0506 + DistanceRo + DistanceRi
reduced w tree:
r_d ~ prp0607 + MaxT0506 + MinT0506 + DistanceRo + DistanceRi + SpRasterNA
Tree only:
r_d ~ SpRasterNA
biological models:
r_d ~ MaxT0506 + pcp0607 + SpRasterNA
r_d ~ MaxT0506 + pcp0607

```{r}
m_1
m_2 <- glm(r_d ~ SpRasterNA - 1, data = bw_env_f, family = binomial(link = "logit"))
m_3 <- glm(r_d ~ MaxT0506 + pcp0607 - 1, data = bw_env_f, family = binomial(link = "logit"))
m_4 <- glm(r_d ~ MaxT0506 + pcp0607 + SpRasterNA - 1, data = bw_env_f, family = binomial(link = "logit"))
m_5 <- glm(r_d ~ pcp0607 + MaxT0506 + MinT0506 + DistanceRo + DistanceRi - 1, data = bw_env_f, family = binomial(link = "logit"))
m_6 <- glm(r_d ~ pcp0607 + MaxT0506 + MinT0506 + DistanceRo + DistanceRi + SpRasterNA - 1, data = bw_env_f, family = binomial(link = "logit"))

BW_models <- list()
BW_models[["m_1"]] <- m_1
BW_models[["m_2"]] <- m_2
BW_models[["m_3"]] <- m_3
BW_models[["m_4"]] <- m_4
BW_models[["m_5"]] <- m_5
BW_models[["m_6"]] <- m_6

modname_BW_models <- paste(names(BW_models))

# get table
library(AICcmodavg)
print(aictab(cand.set = BW_models, modnames = modname_BW_models, sort = TRUE), digits = 3, LL = TRUE)

summary(m_1)
summary(m_5)
summary(m_6)
summary(m_2)
```

# GAM
```{r}
# trial with reduced model
gm_2 <- gam(r_d ~ SpRasterNA + s(X, Y) -1, data = bw_env_xy, family = binomial)
summary(gm_2)     # smooth term significant 

# full model
gm_1 <- gam(r_d ~ MaxT0506 + MinT0506 + pcp0607 + sg__gdd_p1 + DistanceRo + DistanceRi + SpRasterNA + s(X, Y) - 1, data = bw_env_xy, family = binomial(link = "logit"))

# check residuals
sim_resg <- simulateResiduals(fittedModel = gm_1, plot = F)


plot(sim_resg, quantreg=T)    #better
jpeg("QQ-plot gam.jpg", width = 11, height =11, units = "cm", res = 1200, pointsize = 12)
plotQQunif(simulationOutput = sim_resg, 
           testDispersion = FALSE,
           testUniformity = FALSE,
           testOutliers = FALSE)
dev.off()

jpeg("res-fitted gam plot.jpg", width = 11, height =11, units = "cm", res = 1200, pointsize = 12)
plotResiduals(sim_resg, quantreg = T)
dev.off()

# plot against explain variables
plotResiduals(sim_resg, bw_env_f$sg__gdd_p1)
# sg_gdd_p1 good
plotResiduals(sim_resg, bw_env_f$pcp0607)
# pcp0607 good
plotResiduals(sim_resg, bw_env_f$MinT0506)
# MinT0708 quite bad
plotResiduals(sim_resg, bw_env_f$MaxT0506)
# MaxT0506 good
plotResiduals(sim_resg, bw_env_f$DistanceRo)
# DistanceRo not great fror 0.5 and 0.75
plotResiduals(sim_resg, bw_env_f$DistanceRi)
# DistanceRi good
plotResiduals(sim_resg, bw_env_f$SpRasterNA, asFactor = T)

acf(gm_2$residuals, lag.max = 5000)
# lowered the correlation from ~0.8 to ~0.6

jpeg("lagplot gam.jpg", width = 11, height =10, units = "cm", res = 1200, pointsize = 12)
acf(gm_1$residuals, lag.max = 5000, main = " ")
dev.off()

```

# run GAMs
1: full model
2: only tree speceis
3: only previous biologically relevant variables
4: previous biologically relevant variables and tree species
5: reduced model based on significant factors
6: reduced model based on significant factors and tree species
```{r}
gm_3 <- gam(r_d ~ MaxT0506 + pcp0607 + s(X, Y) - 1, data = bw_env_xy, family = binomial(link = "logit"))
gm_4 <- gam(r_d ~ MaxT0506 + pcp0607 + SpRasterNA + s(X, Y) - 1, data = bw_env_xy, family = binomial(link = "logit"))
gm_5 <- gam(r_d ~ pcp0607 + MaxT0506 + DistanceRo + DistanceRi + s(X, Y) - 1, data = bw_env_xy, family = binomial(link = "logit"))
gm_6 <- gam(r_d ~ pcp0607 + MaxT0506 + DistanceRo + DistanceRi + SpRasterNA + s(X, Y) - 1, data = bw_env_xy, family = binomial(link = "logit"))

AIC(gm_1, gm_2, gm_3, gm_4, gm_5, gm_6)

summary(gm_1)
summary(gm_2)
summary(gm_3)
summary(gm_4)
summary(gm_6)
summary(gm_5)
```