---
title: "BW_project"
output: html_document
---

---
title: "BW_random"
output: html_document
---

# load packages and data
```{r}
pacman::p_load(
  tidyverse,
  readxl,
  terra,
  stats,
  dismo,
  usdm,
  corrplot,
  PerformanceAnalytics,
  oddsratio,
  DHARMa,
  MASS,
  nlme,
  mgcv,
  mgcViz)

setwd("~/Documents/Master's/Quantitative Biology/BW_proj")
getwd()
env_r_all_na <- read_excel(path = "./Zhang2022RasterDataLayers/random_pt_xy.xlsx")
env_d_all_na <- read_excel(path = "./Zhang2022RasterDataLayers/defol_pt_xy.xlsx")
```

trying to retreive raster from source file
```{r eval=FALSE, include=FALSE}
aspect_rast <- rast("./Zhang2022RasterDataLayers/aspectRaster.tif")
plot(aspect_rast, main='SpatRaster from file')
```


# prep data
### remove NAs
```{r echo=TRUE}
env_r_all <- env_r_all_na[complete.cases(env_r_all_na), ]
env_d_all <- env_d_all_na[complete.cases(env_d_all_na), ]
```

### create same number of samples for each data set
chose 600 bc largest common centennial 
```{r}
env_r <- env_r_all[sample(nrow(env_r_all), 2400), ]
env_d <- env_d_all[sample(nrow(env_d_all), 2400), ]
```

### combine onto one dataframe
```{r}
# make column specifying defoliation points
env_r$r_d <- c(rep(1, length(env_r$pcp0708)))

# make column specifying random points
env_d$r_d <- c(rep(0, length(env_d$pcp0708)))


# combine and drop columns not common to both and environmental variables not using
bw_env <- rbind(env_d[intersect(names(env_d),names(env_r))],env_r[intersect(names(env_d),names(env_r))])
bw_env <- bw_env[, -c(3, 5, 18, 21)]
# make tree type (SpRasterNA) categorical
bw_env$SpRasterNA <- as.factor(bw_env$SpRasterNA)
```

### subsetted several times?
```{r eval=FALSE, include=FALSE}
# first make env_x_all into one data frame
## make column specifying defoliation points
env_r_all$r_d <- c(rep("r", length(env_r_all$id)))

## make column specifying random points
env_d_all$r_d <- c(rep("d", length(env_d_all$Year)))

## combine and drop columns not common to both
bw_env_all <- rbind(env_d_all[intersect(names(env_d_all),names(env_r_all))],env_r_all[intersect(names(env_d_all),names(env_r_all))])
bw_env_all$SpRasterNA <- as.factor(bw_env_all$SpRasterNA)
bw_env_all <- bw_env_all[, -c(3, 5, 18, 21)]

#then subset repeatedly
iter <- 10
bw_env_rep <- lapply(1:iter, function(i) bw_env_all[sample(nrow(bw_env_all), 600), ])
```

# sample plots
```{r eval=FALSE, include=FALSE}
for(i in seq_along(bw_env[, -19])){
  y <- bw_env[[i]]
plot_env <- ggplot(bw_env, mapping = aes(r_d, y)) + 
  geom_boxplot() +
  ylab(colnames(bw_env)[i])
print(plot_env)
}
```


# correlation analysis
```{r}
##check Multicollinearity - between 14 predictors
predictors_num <- bw_env[, -c(3,18)]
# correlation structure
varCor <- cor(predictors_num)
corrplot(varCor,  tl.col='black')

chart.Correlation(varCor, histogram=F, pch=19)
# based on chart sg_date cor w all temps; sg__gdd_p2 cor w maxTs, pcps and distri; pcps cor w other pcps and ~ max temps (especially 0708 and 0607 w maxt0708 and 0607); minT cor w minTs and ~ maxTs (especially 0607); maxT cor w maxTs and ~ distri; distri cor w sg__gdd_p2 and maxTs

# clustering with dendrogram
allDistNew <- abs(as.dist(cor(predictors_num)))
allClusNew <- hclust(1 - allDistNew)
plot(allClusNew, hang=-1)
# based on diagram: maxTs w other maxTs and sg__gdd_p2, pcps with other pcps, minT w other minT and sg_date

## also so interesting that maxT is correlated somewhat with DistanceRi!!
```

### variance inflation factor
```{r}
# Variance Inflation Factor
## don't really understand how it works and I am happy looking at dendogram and chart and maybe just the initial vif numbers
usdm::vif(as.data.frame(predictors_num))
## distance to rivers and roads fine, sg__gdd_p1 fine, pcp0506 and 0708 relatively fine
## worst sg_date, maxTs, minTs0607
## takes out the highest (sees above 10 as collinearity problem) so that values decrease
usdm::vifstep(as.data.frame(predictors_num))
vifcor(as.data.frame(predictors_num), th=0.8) 
```

### remove variables
```{r}
# my colinearlities are different than Bo's bc I have much smaller/different subset
# still keep MaxT0506 and pcp0607 bc biologically relevant
## remove all but one MaxT (0506), remove sg_date and sg__gdd_p2, minT0607 and minT0708
## this way no overlap in temperature in time may help with lessening colinearity
predictors_num_sel <- predictors_num[,-c(3,4,9,10,12,13)]

# retry
varCor_sel <- cor(predictors_num_sel)
corrplot(varCor_sel,  tl.col='black')
chart.Correlation(varCor_sel, histogram=F, pch=19)

allDistNew_sel <- abs(as.dist(cor(predictors_num_sel)))
allClusNew_sel <- hclust(1 - allDistNew_sel)
plot(allClusNew_sel, hang=-1)


usdm::vif(as.data.frame(predictors_num_sel))
## pcps still very correlated so maybe remove next highest which is pcp0708 and is correlated with X, also remove pcp0506 bc correlated with Y, keeping pcp0607 makes sense beyond biological pretense because the least with lat/long

# retry #2
predictors_num_sel2 <- predictors_num_sel[,-c(4,6)]

varCor_sel2 <- cor(predictors_num_sel2)
corrplot(varCor_sel2, tl.col = "black")
chart.Correlation(varCor_sel2, histogram=F, pch=19)

allDistNew_sel2 <- abs(as.dist(cor(predictors_num_sel2)))
allClusNew_sel2 <- hclust(1 - allDistNew_sel2)
plot(allClusNew_sel2, hang=-1)

usdm::vif(as.data.frame(predictors_num_sel2))
#removing pcp0708 brought all vif under 10

pred_final <- predictors_num_sel2
```

### make data set with only final environmental predictors
without x,y coord
```{r eval=FALSE, include=FALSE}
final_names_e <- c(names(pred_final), "r_d")
str(final_names_e)
bw_env_ef <- bw_env[ ,names(bw_env) %in% final_names_e]
```



model:
full: r_d ~ MaxT0506 + MinT0708 + Prcp0506 + Prcp0607 + Sg__gdd_p1 + DistanceRo + DistanceRi

#### trial fit for glm env
not spatially autocorrelated
```{r eval=FALSE, include=FALSE}
me1 <- glm(r_d ~ ., data = bw_env_ef, family = binomial(link = "logit"))
summary(me1)
coef(me1)
exp(coef(me1))
plot(me1)

me2 <- glm(r_d ~ DistanceRo+pcp0506, data = bw_env_f, family = binomial(link = "logit"))
summary(me2)
exp(coef(me2))
or_glm(data = bw_env_f, model = me2, 
       incr = list(DistanceRo = 4, pcp0506 = 4))
```

## add tree species
```{r}
final_names <- c(names(pred_final), "r_d", "SpRasterNA")
str(final_names)
bw_env_f <- bw_env[ ,names(bw_env) %in% final_names]
```

with x,y coord
```{r}
final_names_xy <- c(names(pred_final), "r_d", "SpRasterNA", "X", "Y")
str(final_names_xy)
bw_env_xy <- bw_env[ ,names(bw_env) %in% final_names_xy]
```

model:
full: r_d ~ MaxT0506 + MinT0506 + Prcp0506 + Prcp0607 + Sg__gdd_p1 + DistanceRo + DistanceRi + SpRasterNA

### trial fit for glm
```{r echo=TRUE}
# tests pairwise comp against intercept (sp1)
m1 <- glm(r_d ~ ., data = bw_env_f, family = binomial(link = "logit"))
summary(m1)
summary(m1)
coef(m1)
exp(coef(m1))
plot(m1)

# test wether each coefficient is diff than 0 **use this
m_1 <- glm(r_d ~  MaxT0506 + MinT0506 + pcp0607 + sg__gdd_p1 + DistanceRo + DistanceRi + SpRasterNA - 1, data = bw_env_f, family = binomial(link = "logit"))
summary(m_1)
summary(fitted(m_1))
coef(m_1)
exp(coef(m_1))
plot(m_1)
## checked if type 3 - ctrl for each level of variable not just variable above


 m_3 <- glm(r_d ~  MinT0506 + pcp0607 +MaxT0506 - 1, data = bw_env_f, family = binomial(link = "logit"))
summary(m_3)
 m_4 <- glm(r_d ~ MaxT0506+ pcp0607 +MinT0506 - 1, data = bw_env_f, family = binomial(link = "logit"))
 summary(m_4)
```

```{r}
# very reduced model to try predict() function
m2 <- glm(r_d ~ DistanceRo, data = bw_env_f, family = binomial(link = "logit"))
summary(m2)
exp(coef(m2))
or_glm(data = bw_env_f, model = m2, 
       incr = list(DistanceRo = 4))


new_pred <- data.frame(DistanceRo = seq(1, 40000, 40))
# try predict again
pred1 <- predict.glm(m2, newdata = new_pred, type = "response")
plot(x = bw_env_f$DistanceRo, y = bw_env_f$r_d)
lines(new_pred$DistanceRo, pred1)
```

#### create test variables
needs work
```{r eval=FALSE, include=FALSE}
env_remaining <- anti_join(env_r_all, env_r)
env_test <- env_remaining[sample(nrow(env_remaining), 10), ]
env_test <- env_test[, -c(1,16)]
```

#### predict with test variables
```{r eval=FALSE, include=FALSE}
pred1 <- predict.glm(m_1, newdata = env_test, type = "response")
plot(y = bw_env_f$r_d)
lines(env_test, pred1)
```

## plot residuals against fitted
about DHARMa:
"The ‘DHARMa’ package uses a simulation-based approach to create readily interpretable scaled (quantile) residuals for fitted (generalized) linear mixed models.
The resulting residuals are standardized to values between 0 and 1 and can be interpreted as intuitively as residuals from a linear regression. The package also provides a number of plot and test functions for typical model misspecification problems, such as over/underdispersion, zero-inflation, and residual spatial and temporal autocorrelation.

To interpret the residuals, remember that a scaled residual value of 0.5 means that half of the simulated data are higher than the observed value, and half of them lower. A value of 0.99 would mean that nearly all simulated data are lower than the observed value. The minimum/maximum values for the residuals are 0 and 1. For a correctly specified model we would expect asymptotically a uniform (flat) distribution of the scaled residuals & uniformity in y direction if we plot against any predictor."
(Hartig 2022)
```{r}
plot(m_1)
# this is very weird and people seem to avoid these sorts of plots for Gzlm
# so use DHARMa package: https://cran.r-project.org/web/packages/DHARMa/DHARMa.pdf and https://cran.r-project.org/web/packages/DHARMa/vignettes/DHARMa.html#workflow-in-dharma
sim_res <- simulateResiduals(fittedModel = m_1, plot = F)

plot(sim_res, quantreg=T)
# definiety normal along Q-Q plot
#red quantile lines mean significant deviation from straight line - outcome depends highly on sample so increase sample size and do repeated measures (MVMC?)
outliers(sim_res)

# test to see where this is coming from
plotResiduals(sim_res, bw_env_f$sg__gdd_p1)
# sg_gdd_p1 good
plotResiduals(sim_res, bw_env_f$pcp0607)
# pcp0607 good
plotResiduals(sim_res, bw_env_f$MinT0506)
# MinT0708 quite bad
plotResiduals(sim_res, bw_env_f$MaxT0506)
# MaxT0506 good
plotResiduals(sim_res, bw_env_f$DistanceRo)
# DistanceRo not great fror 0.5 and 0.75
plotResiduals(sim_res, bw_env_f$DistanceRi)
# DistanceRi good
plotResiduals(sim_res, bw_env_f$SpRasterNA, asFactor = T)
# SpRasterNA good

# how to fix: pcp0506, MinT0708, DistanceRo
# maybe autocorrelation?
testSpatialAutocorrelation(simulationOutput = sim_res, x = bw_env_xy$X, y= bw_env_xy$Y)
# p values is soooooo small lol
# plot against locations
plotResiduals(sim_res, bw_env_xy$X)
plotResiduals(sim_res, bw_env_xy$Y)

# the plots say that the residuals aren't homogenous by test but I think they all look quite straight *I think the power of my sample size is so large my type ii error is huge when the tests being done by the package are keeping the same cut off points

lag.plot(m_1$residuals, lags = 100)
acf(m_1$residuals, lag.max = 1700)
## at a large scale it seems there is some pattern in the residuals still (likely spatial autocorrelation)


# found the reason for the weird lag plot by making the plot bigger
acf(bw_env_xy$pcp0607, lag.max = 5000)
acf(bw_env_f$MaxT0506, lag.max = 5000)
acf(bw_env_xy$DistanceRo, lag.max = 5000)

# ok
acf(bw_env_xy$DistanceRi, lag.max = 5000)
acf(bw_env_xy$MinT0506, lag.max = 5000)

# these are fine
acf(bw_env_xy$sg__gdd_p1, lag.max = 5000)
acf(bw_env_xy$X, lag.max = 5000)
acf(bw_env_xy$Y, lag.max = 5000)

```

# models to test
full model:
r_d ~ MaxT0506 + MinT0506 + pcp0607 + Sg__gdd_p1 + DistanceRo + DistanceRi + SpRasterNA
reduced model:
r_d ~ prp0607 + MaxT0506 + MinT0506 + DistanceRo + DistanceRi
reduced w tree:
r_d ~ prp0607 + MaxT0506 + MinT0506 + DistanceRo + DistanceRi + SpRasterNA
Tree only:
r_d ~ SpRasterNA
biological models:
r_d ~ MaxT0506 + pcp0607 + SpRasterNA
r_d ~ MaxT0506 + pcp0607

```{r}
m_1
m_2 <- glm(r_d ~ SpRasterNA - 1, data = bw_env_f, family = binomial(link = "logit"))
m_3 <- glm(r_d ~ MaxT0506 + pcp0607 - 1, data = bw_env_f, family = binomial(link = "logit"))
m_4 <- glm(r_d ~ MaxT0506 + pcp0607 + SpRasterNA - 1, data = bw_env_f, family = binomial(link = "logit"))
m_5 <- glm(r_d ~ pcp0607 + MaxT0506 + MinT0506 + DistanceRo + DistanceRi - 1, data = bw_env_f, family = binomial(link = "logit"))
m_6 <- glm(r_d ~ pcp0607 + MaxT0506 + MinT0506 + DistanceRo + DistanceRi + SpRasterNA - 1, data = bw_env_f, family = binomial(link = "logit"))

BW_models <- list()
BW_models[["m_1"]] <- m_1
BW_models[["m_2"]] <- m_2
BW_models[["m_3"]] <- m_3
BW_models[["m_4"]] <- m_4
BW_models[["m_5"]] <- m_5
BW_models[["m_6"]] <- m_6

modname_BW_models <- paste(names(BW_models))

# get table
library(AICcmodavg)
print(aictab(cand.set = BW_models, modnames = modname_BW_models, sort = TRUE), digits = 3, LL = TRUE)

summary(m_1)
```

taking out insignificant variables does not reduce likelihood substantially
but taking out tree species does
only using tree species or previously understood environmental variables is not nearly as likely (or combined)


# try a gam
```{r}
gm_2 <- gam(r_d ~ SpRasterNA + s(X, Y) -1, data = bw_env_xy, family = binomial)
summary(gm_2)

# that looks great and smooth term veeeery significant (like what was found for DAHRMa)
gm_2x <- gam(r_d ~ SpRasterNA + s(X) -1, data = bw_env_xy, family = binomial)
summary(gm_2x)
gm_2y <- gam(r_d ~ SpRasterNA + s(Y) -1, data = bw_env_xy, family = binomial)
summary(gm_2y)

gm_1 <- gam(r_d ~ MaxT0506 + MinT0506 + pcp0607 + sg__gdd_p1 + DistanceRo + DistanceRi + SpRasterNA + s(X, Y) - 1, data = bw_env_xy, family = binomial(link = "logit"))

sim_resg <- simulateResiduals(fittedModel = gm_1, plot = F)

plot(sim_resg, quantreg=T)

# looks soooo much better

# plot against explain variables
plotResiduals(sim_resg, bw_env_f$sg__gdd_p1)
# sg_gdd_p1 good
plotResiduals(sim_resg, bw_env_f$pcp0607)
# pcp0607 good
plotResiduals(sim_resg, bw_env_f$MinT0506)
# MinT0708 quite bad
plotResiduals(sim_resg, bw_env_f$MaxT0506)
# MaxT0506 good
plotResiduals(sim_resg, bw_env_f$DistanceRo)
# DistanceRo not great fror 0.5 and 0.75
plotResiduals(sim_resg, bw_env_f$DistanceRi)
# DistanceRi good
plotResiduals(sim_resg, bw_env_f$SpRasterNA, asFactor = T)

testSpatialAutocorrelation(simulationOutput = sim_resg, x = bw_env_xy$X, y= bw_env_xy$Y)
# p values is soooooo small lol
# plot against locations
plotResiduals(sim_resg, bw_env_xy$X)
plotResiduals(sim_resg, bw_env_xy$Y)

# the plots say that the residuals aren't homogenous by test but I think they all look quite straight *I think the power of my sample size is so large my type ii error is huge when the tests being done by the package are keeping the same cut off points

lag.plot(gm_2$residuals, lags = 1)
acf(gm_2$residuals, lag.max = 1700)
# lowered the correlation from ~0.8 to ~0.6
## at a large scale it seems there is some pattern in the residuals still (likely spatial autocorrelation)

```

# run GAMs
```{r}
gm_3 <- gam(r_d ~ MaxT0506 + pcp0607 + s(X, Y) - 1, data = bw_env_xy, family = binomial(link = "logit"))
gm_4 <- gam(r_d ~ MaxT0506 + pcp0607 + SpRasterNA + s(X, Y) - 1, data = bw_env_xy, family = binomial(link = "logit"))
gm_5 <- gam(r_d ~ pcp0607 + MaxT0506 + MinT0506 + DistanceRo + DistanceRi + s(X, Y) - 1, data = bw_env_xy, family = binomial(link = "logit"))
gm_6 <- gam(r_d ~ pcp0607 + MaxT0506 + DistanceRo + DistanceRi + SpRasterNA + s(X, Y) - 1, data = bw_env_xy, family = binomial(link = "logit"))

AIC(gm_1, gm_2, gm_3, gm_4, gm_5, gm_6)

summary(gm_1)
summary(gm_6)
```


